Popularity Adjusted Block Models are Generalized Random Dot Product Graphs

### 2022-03-28 fls-lightning-talk/slides.Rmd
updated script and minor tweaks to slides
changes based on feedback

How can we cluster the nodes of a network?

Statistical inference (parametric approach):

1. Define a generative model for graph: 
$G \mid z_1, ..., z_n, \vec{\theta} \sim P(\vec{z}, \vec{\theta})$.
2. Develop a method for obtaining estimators: 
$f(G) = (\hat{\vec{z}}, \hat{\vec{\theta}})$.
3. Describe asymptotic properties of estimators: 
$(\hat{\vec{z}}, \hat{\vec{\theta}}) \to (\vec{z}, \vec{\theta})$.

@@

:::: {.column width=62.5%}

@@

:::: {.column width=37.5%}

**Example 1**: If every entry $P_{ij} = \theta$, then $A \sim \BG(P)$ is an Erdos-Renyi graph.  
For this model, $A_{ij} \iid \Bernoulli(\theta)$.

```{r, fig.height = 2, fig.width = 2}
n <- 2 ** 5
p <- 1 / log(n)
P <- matrix(p, nrow = n, ncol = n)
A <- draw.graph(P)
qgraph::qgraph(A, vsize = 4)
```

@@

**Example 2**: Stochastic Block Model with two communities

@@

* To make this an assortative SBM, set $p q > r^2$.
* In this example, $p = 1/2$, $q = 1/4$, and $r = 1/8$.

@@

Let each vertex $i \in [n]$ have $K$ popularity parameters
$\lambda_{i1}, ..., \lambda_{iK} \in [0, 1]$. 
Then $A \sim \PABM(\{\lambda_{ik}\}_K)$ if each $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$,  
e.g., if $z_i = k$ and $z_j = l$, $P_{ij} = \lambda_{il} \lambda_{jk}$.

@@

$A \sim \GRDPG_{p, q}(X)$ iff

* Latent vectors $x_1, ..., x_n \in \mathbb{R}^{p+q}$ such that 
$x_i^\top I_{p, q} x_j \in [0, 1]$ and $I_{p, q} = \blockdiag(I_p, -I_q)$
* $A \sim \BG(X I_{p, q} X^\top)$ where 
$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$
  * $P(\text{edge between } v_i, v_j) = x_i^\top I_{p,q} x_j$

## (Generalized) Random Dot Product Graph Model

### Recovery/Estimation

Want to estimate $X$ from $A$, or alternatively, 
interpoint distances, inner products, or angles.

### Adjacency Spectral Embedding

To embed in $\mathbb{R}^{p+q}$,

1. Compute $A \approx \hat{V} \hat{\Lambda} \hat{V}^\top$ 
where $\hat{\Lambda} \in \mathbb{R}^{(p+q) \times (p+q)}$ and 
$\hat{V} \in \mathbb{R}^{n \times (p+q)}$ 
by using the $p$ most positive and $q$ most negative eigenvalues.

2. Let $\hat{X} = \hat{V} |\hat{\Lambda}|^{1/2}$.

\vspace*{.5\baselineskip}

$\max\limits_i \|\hat{X}_i - Q_n X_i \| = O_P \Big( \frac{(\log n)^c}{n^{1/2}} \Big)$
(Rubin-Delanchy et al., 2020)

@@

**Example 2** (cont'd): Assortative SBM ($p q > r^2$) with $K = 2$

@@

**Example 2** (cont'd): If we want to perform community detection,

@@

$$A \sim \PABM(\{\lambda^{(kl)}\}_K) \iff A \sim \GRDPG_{p, q}(X U)$$

@@

## Conclusion

1. The PABM is a recently developed flexible block model that can be used to describe graphs with community structure.

2. The GRDPG, which can describe all block models, motivates a spectral approach to statistical inference on graphs.

3. Under the GRDPG framework, the PABM with $K$ communities can be induced by a latent configuration in $\mathbb{R}^{K^2}$ consisting of $K$ $K$-dimensional subspaces that are orthogonal to each other.

4. The latent configuration of the PABM under the GRDPG framework leads to an intuitive method for community detection with nice theoretical asymptotic properties.

>>>>

How might we cluster the nodes of a network?

@@

@@

@@

**Example 1**: Stochastic Block Model with $K=2$ communities

@@

@@

Let each vertex $i \in [n]$ have $K$ popularity parameters
$\lambda_{i1}, ..., \lambda_{iK} \in [0, 1]$. 
Then $A \sim \PABM(\{\lambda_{ik}\}_K)$ if each $P_{ij} = \lambda_{i z_j} \lambda_{j z_i}$.

@@

Let $I_{p,q} = \blockdiag(I_p, -I_q)$ and suppose that $x_1, \ldots, x_n \in \mathbb{R}^{p+q}$ are such that $x_i^\top I_{p,q} x_j \in [0,1]$.  
Then $A \sim \GRDPG_{p, q}(X)$ iff $A \sim \BG(X I_{p,q} X^\top)$, where $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$

Adjacency Spectral Embedding (Sussman et al., 2012) estimates $x_1, ..., x_n \in \mathbb{R}^{p+q}$ from $A$:

1. Let $\hat{\Lambda}$ be the diagonal matrix that contains the absolute values of the $p$ most positive and the $q$ most negative eigenvalues.
2. Let $\hat{V}$ be the matrix whose columns are the corresponding eigenvectors.
3. Compute $\hat{X} = \hat{V} \hat{\Lambda}^{1/2}$.

Theorem: $\max\limits_i \|\hat{X}_i - Q_n X_i \| = O_P \Big( \frac{(\log n)^c}{n^{1/2}} \Big)$

@@

**Example 1** (cont'd): Assortative SBM ($p q > r^2$) with $K = 2$

@@

**Example 1** (cont'd): If we want to perform community detection,

@@

$$A \sim \PABM(\{\lambda_{ik}\}_K) \iff A \sim \GRDPG_{p, q}(X U)$$

@@
